Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-07-19 paper.Rnw
References fixes, author affiliations, figure fixes

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.

@@

% add \subsubsubsection
% \setcounter{secnumdepth}{4}  
% \titleformat{\paragraph}
% {\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
% \titlespacing*{\paragraph}
% {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
%notes and TODO formatting
%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}
%\SweaveOpts{concordance=TRUE}

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author{Pipitone J., Park M., Winterburn J., Lett, T., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\todo{Full names}
\todo{Associations}
\maketitle

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\todo[inline]{finish}

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007,Weiner2013}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation.  In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods.  The resulting nonlinear
transformation is applied to the manual labels (ie. {\em label propagation}) to
apply them to the target image.  While this methodology has been used
successfully in several contexts
\citep{MallarMedIA200809,MallarHBM20082009,Collins1995,Haller1997}, it is limited
in accuracy by the introduction of errors due inaccuracies in the nonlinear
transformation itself, partial volume effects in label resampling, and
irreconcilable differences between the neuroanatomy represented within the atlas
and target images.

@@

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty06a,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data \citep{Chakravarty2013} is possible.  The
novelty of this manuscript is the extension our multi-atlas methodology to the
hippocampus using more than a single input atlas, while simultaneously limiting
the number of possible inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

@@

\ants is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation \citep{Avants}. The transformation is estimated in a
hierarchical fashion where the MRI data is subsampled, allowing large
deformations to be estimated and successively refined at later hierarchical
stages (where the data is subsampled to a finer grid). The deformation field and
the objective function are regularized with a Gaussian kernel at each level of
the hierarchy. The \ants algorithm is freely available
\url{http://www.picsl.upenn.edu/ANTS/}. We used an implementation of the \ants
algorithm compatible with the MINC data format, mincANTS
\url{https://github.com/vfonov/mincANTS}.

@@

  \item[Cross-correlation Weighted Majority Vote]
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{Chakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter and displayed as the size of the template library
  in the rest of the paper.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-COMPLETE
% Prepares the data in a the form needed for plotting.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<ADNI-src-volumes-prep,echo=F,cache=T>>=
totals.complete  <- read.csv('data/cache/ADNI1:totals.complete.csv')
totals           <- read.csv('data/cache/ADNI1:totals.csv')
package_totals   <- read.csv('data/cache/ADNI1:package_totals.csv')
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                             
% ADNI1 Complete 1Yr Dataset Demographics Table 
%                                                                             
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]
yr1 = 
adnimerge.yr1 = merge(adnimerge, yr1) 
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge.yr1, 
               #subset = RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-prep,dependson='ADNI1-xval-load',include=FALSE,cache=TRUE>>=
all_data  <- read.csv('data/cache/ADNI-XVAL:all_data.csv')
all_data_mean  <- read.csv('data/cache/ADNI-XVAL:all_data_mean.csv')
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/subfield-montage.pdf}
  \end{centering}
  \caption{Sagittal slices from two subjects showing a comparison of the
  original Winterburn atlas subfield segmentations (at 0.3mm3-isotropic voxel
  resolution), the subsampled Winterburn segmentations (at 0.9mm3-isotropic
  voxel resolution), and the \mb labels on the subsampled atlas image.}
  \label{fig:subfield-montage}
  \todo{MAGeT on BRAVO.  Also, show coronal views}
\end{figure}

@@

casted = dcast(mean_sz_vols, Subject + Label ~ Method)
lm = lm(MAGeT ~ Manual, casted)
ggplot(data=casted, aes(x = Manual, y = MAGeT)) + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_point() + 
  geom_text(aes(x=3500, y=5000, label=lm_eqn(lm),hjust=0,size=1), parse=TRUE, data=data.frame()) +
  xlab(expression("Mean Manual hippocampus volume " (mm^3))) + 
  ylab(expression("Mean MAGeT hippocampus volume "  (mm^3))) + 
  scale_size_continuous(guide="none") 
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volume for \mb against manually rated Hippocampal volumes}
  \label{fig:SZ-volumes}
\end{figure}

@@

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly
chosen subject images (7 from each disease class) and used majority vote label
fusion. The entire \adnidataset dataset was segmented by \mb and we 
now compare the resulting volumes with those obtained by manual segmentation (SNT), 
and other automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADN1-1.5T-Complete-1yr-Dataset-Segmentations} shows the total count of 
segmentations available, including a count of those which have failed a quality 
control inspection. Only those images which had segmentations from every method
are included in the following analysis (a total of
\Sexpr{length(totals.complete$RID)} images; Table 
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals}).

We find a close relationship in total bilateral hippocampal volume between
all methods and manually segmented volumes (Figure \ref{fig:ADNI-scr-volumes-plot}). 
Volumes are correlated with Pearson-r > 0.78 for all methods across disease categories.
Within disease categories (Figure \ref{fig:ADNI-scr-volumes-boxplot}),
\mb is consistently well correlated to manual volumes (Pearson-r > 0.85), but appears to 
slightly over-estimate the volume of the AD hippocampus.

To investigate the level of agreement with manually segmented hippocampal volumes, 
we constructed Bland-Altman plots for each method (Figure \ref{fig:ADNI1-src-Bland-Altman}).  
As Bland \& Altman, 1985 \cite{Bland1986} noted, high 
correlation amongst measures of the same quantity does not necessarily 
imply agreement (as correlation can be driven by a large range in true values, 
for instance).  What is most striking in Figure \ref{fig:ADNI1-src-Bland-Altman}
is that all methods show an obvious proportional bias: FreeSurfer and FSL 
markedly under-estimate smaller hippocampi and over-estimate large hippocampi, 
whereas MAPER and \mb more conservatively show the reverse bias.  Additionally, 
all methods show a fixed bias, with FreeSurfer and FSL most dramatically 
over-estimating hippocampal volume by $2600mm^3$ and $2800mm^3$ on average, respectively, 
and MAPER and \mb within $250mm^3$ on average.

Figure \ref{fig:ADNI-scr-subfields} shows a qualitative comparison of \mb and 
manual (SNT) hippocampal segmentations for 10 randomly selected subjects in 
each disease category, and illustrates some of the common errors found during 
visual inspection. Mostly frequently, we find \mb improperly includes the vestigial
hippocampal sulcus and, although not anatomically incorrect, \mb under-estimates 
the hippocampal body in comparison to the manual (SNT) segmentation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 1yr Complete Dataset Package Totals & Failures
%                        
<<ADNI1-scr-package-totals, echo=F, results="asis",cache=FALSE>>=
caption = paste(
  "Pass/fail quality control indicators were supplied with the FreeSurfer",
  "volumes downloaded from the ADNI website (we used the temporal lobe quality",
  "control indicator, TEMPQC). One of the authors (MP) performed visual",
  "quality inspection for MAGeT and FSL segmentations.", sep=" ")
latex(package_totals, file="", size="scriptsize", 
    caption = caption,
    title = "",
    label="tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[h]
<<ADNI-scr-volumes-boxplot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=7,fig.height=4>>=
melted = melt(totals.complete,measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT"),
              variable.name="Method", value.name="Volume")
correlations = ddply(melted, c("DX","Method"), function (df) {
  data.frame(
    pearson = cor(df$SNT, df$Volume)
  )
})
m2 = melt(totals.complete,measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT","SNT"), variable.name="Method")
levels(m2$Method) = c("FreeSurfer"= "FreeSurfer", "FSL" = "FSL", "MAPER" = "MAPER", "MAGeT" = "MAGeT", "SNT" = "Manual")
m2$DX = factor(m2$DX, levels(m2$DX)[c(2,3,1)])

qplot(DX,value,data=m2,
      colour=Method,geom="boxplot") + 
      #geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
      #          colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Combined whole hippocampal volume (", mm^3, ")"))) + 
      theme(legend.direction = "horizontal", legend.position = "bottom", text=element_text(size=18))
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT) by disease category.  Pearson correlation 
  with manual volumes are shown below each box-and-whisker.}
  \label{fig:ADNI-scr-volumes-boxplot}
\end{figure}

\begin{figure}[h]
<<ADNI-scr-volumes-plot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=6,fig.height=4>>=
melted=melt(totals.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "Method", value.name = "volume")

df = totals.complete

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point() + 
    geom_smooth(method="lm") + 
    xlab(expression("Manual bilateral whole hippocampal volume " (mm^3))) + 
    ylab(expression("Automated bilateral whole hippocampal volume " (mm^3))) +
    annotate("text", y = c(3000,2500,2000,1500), x=Inf-100, size=3,
              label = c(paste("FS r=", round(cor(df$SNT, df$FS),2)),
                        paste("FSL r=", round(cor(df$SNT, df$FSL),2)),
                        paste("MAPER r=", round(cor(df$SNT, df$MAPER),2)),
                        paste("MAGeT r=", round(cor(df$SNT, df$MAGeT),2))),
              hjust=1.1,vjust=1.5) +
    theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT). Pearson correlation with manual volumes
  are shown for each method.}
  \label{fig:ADNI-scr-volumes-plot}
\end{figure}

\begin{figure}
<<ADN1-src-Bland-Atlman, cache=TRUE, dependson='ADNI-baseline-volumes-prep', fig.width=6, fig.height=6>>=
names(totals.complete)[8] = "FreeSurfer"
melted=melt(totals.complete, measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.
\usepackage{authblk}    % Listing Author affiliations

@@

% styling
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,5]{Aristotle Voineskos}
\author[1,5,6]{M. Mallar Chakravarty}
\author{Alzheimer's Disease Neuroimaging Initiative}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Brain Imaging Group, Douglas Hospital Research Centre, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[6]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}

\renewcommand\Authands{ and }

\begin{document}
\maketitle

\begin{abstract}
%Neuroimaging research often relies on automated anatomical segmentations of MR
%images of the brain. Current multi-atlas based approaches provide accurate
%segmentations of brain images by propagating manually derived segmentations of
%specific neuroanatomical structures to unlabelled data. These approaches often
%rely on a large number of such manually segmented atlases that take significant
%time and expertise to produce. We present an algorithm for the automatic
%segmentation of the hippocampus that minimizes the number of atlases needed
%while still achieving similar accuracy to multi-atlas approaches.

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation.  In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods.  The resulting nonlinear
transformation is applied to the manual labels (ie. {\em label propagation}) to
apply them to the target image.  While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited
in accuracy by the introduction of errors due inaccuracies in the nonlinear
transformation itself, partial volume effects in label resampling, and
irreconcilable differences between the neuroanatomy represented within the atlas
and target images.

@@

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty2006,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data \citep{MallarChakravarty2012} is possible.  The
novelty of this manuscript is the extension our multi-atlas methodology to the
hippocampus using more than a single input atlas, while simultaneously limiting
the number of possible inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

@@

\ants is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation \citep{Avants2008}. The transformation is estimated in a
hierarchical fashion where the MRI data is subsampled, allowing large
deformations to be estimated and successively refined at later hierarchical
stages (where the data is subsampled to a finer grid). The deformation field and
the objective function are regularized with a Gaussian kernel at each level of
the hierarchy. The \ants algorithm is freely available
\url{http://www.picsl.upenn.edu/ANTS/}. We used an implementation of the \ants
algorithm compatible with the MINC data format, mincANTS
\url{https://github.com/vfonov/mincANTS}.

@@

  \item[Cross-correlation Weighted Majority Vote]
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{MallarChakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter and displayed as the size of the template library
  in the rest of the paper.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-COMPLETE
% Prepares the data in a the form needed for plotting.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<ADNI-volumes-prep,echo=F,cache=T>>=
means.complete   <- read.csv(gzfile('data/cache/ADNI1:means.complete.csv.gz'))
package_totals   <- read.csv(gzfile('data/cache/ADNI1:package_totals.csv.gz'))
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                             
% ADNI1 Complete 1Yr Dataset Demographics Table 
%                                                                             
<<ADNI-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]
yr1 = 
adnimerge.yr1 = merge(adnimerge, yr1) 
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge.yr1, 
               #subset = RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-prep,dependson='ADNI1-xval-load',include=FALSE,cache=TRUE>>=
all_data  <- read.csv(gzfile('data/cache/ADNI-XVAL:all_data.csv.gz'))
all_data_mean  <- read.csv(gzfile('data/cache/ADNI-XVAL:all_data_mean.csv.gz'))
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/subfield-montage.pdf}
  \end{centering}
  \caption{Sagittal slices from two subjects showing a comparison of the
  original Winterburn atlas subfield segmentations (at 0.3mm3-isotropic voxel
  resolution), the subsampled Winterburn segmentations (at 0.9mm3-isotropic
  voxel resolution), and the \mb labels on the subsampled atlas image.}
  \label{fig:subfield-montage}
\end{figure}

\todo{MAGeT on BRAVO.  Also, show coronal views}

@@

casted = dcast(mean_sz_vols, Subject ~ Method, value.var = "Volume")
lm = lm(MAGeT ~ Manual, casted)
ggplot(data=casted, aes(x = Manual, y = MAGeT)) + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_point() + 
  geom_text(aes(x=3500, y=5000, label=lm_eqn(lm),hjust=0,size=1), parse=TRUE, data=data.frame()) +
  xlab(expression("Mean Manual hippocampus volume " (mm^3))) + 
  ylab(expression("Mean MAGeT hippocampus volume "  (mm^3))) + 
  scale_size_continuous(guide="none") 
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volume for \mb against manually rated Hippocampal volumes}
  \label{fig:SZ-volumes}
\end{figure}

@@

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly
chosen subject images (7 from each disease class) and used majority vote label
fusion. The entire \adnidataset dataset was segmented by \mb and we 
now compare the resulting volumes with those obtained by manual segmentation (SNT), 
and other automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADN1-1.5T-Complete-1yr-Dataset-Segmentations} shows the total count of 
segmentations available, including a count of those which have failed a quality 
control inspection. Only those images which had segmentations from every method
are included in the following analysis (a total of
\Sexpr{length(means.complete$RID)} images; Table 
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals}).

We find a close relationship in total bilateral hippocampal volume between
all methods and manually segmented volumes (Figure \ref{fig:ADNI-volumes-plot}). 
Volumes are correlated with Pearson-r > 0.78 for all methods across disease categories.
Within disease categories (Figure \ref{fig:ADNI-volumes-boxplot}),
\mb is consistently well correlated to manual volumes (Pearson-r > 0.85), but appears to 
slightly over-estimate the volume of the AD hippocampus.

To investigate the level of agreement with manually segmented hippocampal volumes, 
we constructed Bland-Altman plots for each method (Figure \ref{fig:ADNI1-Bland-Altman}).  
As Bland \& Altman, 1985 \cite{Bland1986} noted, high 
correlation amongst measures of the same quantity does not necessarily 
imply agreement (as correlation can be driven by a large range in true values, 
for instance).  What is most striking in Figure \ref{fig:ADNI1-Bland-Altman}
is that all methods show an obvious proportional bias: FreeSurfer and FSL 
markedly under-estimate smaller hippocampi and over-estimate large hippocampi, 
whereas MAPER and \mb more conservatively show the reverse bias.  Additionally, 
all methods show a fixed bias, with FreeSurfer and FSL most dramatically 
over-estimating hippocampal volume by $2600mm^3$ and $2800mm^3$ on average, respectively, 
and MAPER and \mb within $250mm^3$ on average.

Figure \ref{fig:ADNI-subfields} shows a qualitative comparison of \mb and 
manual (SNT) hippocampal segmentations for 10 randomly selected subjects in 
each disease category, and illustrates some of the common errors found during 
visual inspection. Mostly frequently, we find \mb improperly includes the vestigial
hippocampal sulcus and, although not anatomically incorrect, \mb under-estimates 
the hippocampal body in comparison to the manual (SNT) segmentation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 1yr Complete Dataset Package Totals & Failures
%                        
<<ADNI-seg-package-totals, echo=F, results="asis",cache=FALSE>>=
caption = paste(
  "Pass/fail quality control indicators were supplied with the FreeSurfer",
  "volumes downloaded from the ADNI website (we used the temporal lobe quality",
  "control indicator, TEMPQC). One of the authors (MP) performed visual",
  "quality inspection for MAGeT and FSL segmentations.", sep=" ")
latex(package_totals, file="", size="scriptsize", 
    caption = caption,
    title = "",
    label="tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[h]
<<ADNI-volumes-boxplot,dependson='ADNI-volumes-prep',cache=T,fig.width=7,fig.height=4>>=
melted = melt(means.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT"),
              variable.name="Method", value.name="Volume")
correlations = ddply(melted, c("DX","Method"), function (df) {
  data.frame(
    pearson = cor(df$SNT, df$Volume)
  )
})
m2 = melt(means.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT","SNT"), variable.name="Method")
levels(m2$Method) = c("FS"= "FreeSurfer", "FSL" = "FSL", "MAPER" = "MAPER", "MAGeT" = "MAGeT", "SNT" = "Manual")
m2$DX = factor(m2$DX, levels(m2$DX)[c(2,3,1)])

qplot(DX,value,data=m2,
      colour=Method,geom="boxplot") + 
      #geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
      #          colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Mean hemispheric hippocampal volume (", mm^3, ")"))) + 
      theme(legend.direction = "horizontal", legend.position = "bottom", text=element_text(size=18))
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT) by disease category.  Pearson correlation 
  with manual volumes are shown below each box-and-whisker.}
  \label{fig:ADNI-volumes-boxplot}
\end{figure}

\begin{figure}[h]
<<ADNI-volumes-plot,dependson='ADNI-volumes-prep',cache=T,fig.width=6,fig.height=4>>=
melted=melt(means.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "Method", value.name = "volume")

df = means.complete

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point() + 
    geom_smooth(method="lm") + 
    xlab(expression("Manual mean hemispheric hippocampal volume " (mm^3))) + 
    ylab(expression("Automated mean hemispheric hippocampal volume " (mm^3))) +
#    annotate("text", y = c(3000,2500,2000,1500), x=Inf-100, size=3,
#              label = c(paste("FS r=", round(cor(df$SNT, df$FS),2)),
#                        paste("FSL r=", round(cor(df$SNT, df$FSL),2)),
#                        paste("MAPER r=", round(cor(df$SNT, df$MAPER),2)),
#                        paste("MAGeT r=", round(cor(df$SNT, df$MAGeT),2))),
#              hjust=1.1,vjust=1.5) +
    theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT). Pearson correlation with manual volumes
  are shown for each method.}
  \label{fig:ADNI-volumes-plot}
\end{figure}

\begin{figure}
<<ADN1-Bland-Atlman, cache=TRUE, dependson='ADNI-volumes-prep', fig.width=6, fig.height=6>>=
names(means.complete)[8] = "FreeSurfer"
melted=melt(means.complete, measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")
